---
title: "R markdown"
author: "Simon Wood"
output: pdf_document
---
# various header
## sizes are available 

Using Latex mathematics type-setting, you can include maths inline, $\int f^{\prime\prime}(x)^2 dx$, or displayed
$$
\hat { \beta} = \underset{\beta}{{\rm argmin}} ~ \|{\bf y} - {\bf X}{ \beta }\|^2.
$$


***

Quotes can be added, and words emphasized...

> **Likelihood**. Parameter values that make the observed data probable according to the model are more *likely* to be correct than parameter values that make the data appear improbable.

Furthermore

* lists
  * sub lists (indented 2 spaces)
  * etc
* can be added

R code can be included, inline to produce a `r paste("res","ult",sep="")`, or as a *code chunk* for display, but maybe not running...
```{r eval=FALSE}
paste("res","ult",sep="") ## can comment too
```
... or display and running, as here...
```{r} 
paste("res","ult",sep="")
```
... or running without display, as here...
```{r echo=FALSE} 
paste("res","ult",sep="")
```
Figures are easily added too
```{r fig.width=3,fig.height=2.8,fig.align='center'}
hist(nhtemp,main="")
```